---
title: "Estimation of sample size"
author: "AGBOHOUTO OMRAAM OLIVIER"
date: "2023-06-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("epiDisplay")
library("MASS")
library("survival")
```

## Overview

In clinical trial or epidemiological study, one of the critical point is the estimation of the good sample size which will help to grant detection of significant difference and then get a good power for statistical test in our study.

It is important to have the good sample size for many reason. Firstly, for ethic aspect. Small sample will not allow to detect significant effect or can conduct to wrong conclusion and large sample will include patients or individual that we would not need. Then we can loose human, financial and logistic resources.

Determination of sample size is based on power of statistical test. Power of statistical test is the probability to reject null hypothesis knowing that alternative hypothesis is right. It means that power is the complementary of type II risk denoted $\beta$ :  $P= 1-\beta$.

In other words, power of statistical test is the probability to conclude to a true significative difference on the priincipal outcome. In this work, we will only focus on the case of one simple principal outcome with a statistical test to attend the primary endpoint. We will provide another work in case of composite principal outcome.

Power of statistical test depends on some factors such as:
- Sample size
- Difference to be detected
- Test significance level or type I risk denoted $\alpha$
- Variability of observations ( for example correlation between observations, variance, lost to follow up)

Thus a value of sample size determine a value of power. And a value of power define of value of sample size. Usually, to compute sample size we should:
- define the primary endpoint and the statistical test to be used 
- define the expected difference or value of principal endpoint of the study (for instance difference and standard deviation of mean height between treatment group, proportion of an event in the exposition group, survival duration without disease progression between exposition group)
- define the ratio between exposition group (case control study)
- define lost to follow up percentage if needed
- define the risks $\alpha$ and $\beta$. Usually $\alpha=0.05$ and $\beta=0.8$.

Once all previous parameters defined, there are different ways to compute sample size in different situation or design of study.

- Using formula of sample size: Formula to compute sample size depends on principal outcome (proportion comparison, mean comparison, Odds ratio or relative risk, time-to-event analysis).
 - Proportion comparison: $N=\frac{2 \times \pi (1-\pi) (z_{1-\alpha /2} + z_{1-\beta})^2}{(p_1 - p_2)^2}$ where $p_1$ is the proportion of event in group of exposed patients, $p_2$ is the proportion of event in group of non exposed patients, $\pi = \frac{p_1+p_2}{2}$, $z_{\alpha/2}$ the quantile associated to level $\alpha/2$ and $z_{\beta}$ the quantile associated to level $\beta$. We can notice that N increase when difference between proportion is small.
 - Mean comparison: $N=\frac{2 \times \sigma^2 (z_{1-\alpha /2} + z_{1-\beta})^2}{(\bar{x}_1 - \bar{x}_2)^2}$ where $\bar{x}_1$ is the mean of outcome is group 1, $\bar{x}_2$ is the mean of outcome is group 2, $\sigma$ the standard deviation of outcome, $z_{\alpha/2}$ the quantile associated to level $\alpha/2$ and $z_{\beta}$ the quantile associated to level $\beta$
 - OR of intervention parameter: same formula for proportion comparison with $p_1=\frac{1}{1+\frac{1-p_2}{OR \times p_2}}$. This supposed to know also the baseline value of proportion of event in the control group in addition of all others parameters enumerated in estimation of sample size in case of proportion comparison.
 - RR in longitudinal analysis: same formula for proportion comparison with $p_1$ incidence in exposed group, $p_2$ incidence in non exposed group, $p_1= RR \times p_2$. This supposed to know also the baseline value of proportion of event in the non exposed group in addition of all others parameters enumerated in estimation of sample size in case of proportion comparison.

Note that all these formula do not take into account the ratio of patient distribution (1 case for 4 control for example) or rate of lost to follow up. These two factors can impact size of sample and need to be taken into account. In case of ratio, we introduce a factor $\gamma = \frac{n_2}{n_1}$ which represent the ratio in the previous formula. For example for proportion comparison we have:
- Proportion comparison: $N=\frac{(z_{1-\alpha /2 } \sqrt{(1+\gamma).\pi.(1-\pi)} + z_{1-\beta} \sqrt{\gamma . p_1(1-p_1)+p_2(1-p_2)})^2}{\gamma \times (p_1 - p_2)^2}$


- Using statistical software such as R (package epiDisplay or epiR), SAS (PROC POWER),or PASS: There are many software and webapp which offer sample size calculation. We focus on R with package epiDisplay. With this package we can estimate sample size and power of a study. There are many functions in this package to estimate either sample size, or power in different design of study and for different test of primary outcome.

- Using empirical simulation: The process of this method is simple but you must define a size for your sample in advance, then estimate the power based on many different sample simulate and then select the good sample size with the best power or the sample size from which we have not a significant gain of power. In this method we consider the power as the proportion of significant p-value. It means that we test the primary endpoint on each sample simulated and calculate the proportion of times that we have a significant p-value (p-value < $\alpha$) over all p-values computed.

### Proportion comparison
This example is taken from clinical trial Randomised, double-blind, placebo-controlled trial of interferon alpha-2b with and without ribavirin for chronic hepatitis C. The Swedish Study Group of Lancet. 1998 Jan 10;351(9096):83-7. https://pubmed.ncbi.nlm.nih.gov/9439491/

In this study we want to prove the efficacy of the combination of interferon alpha-2b and ribavirin in treatment of chronic hepatitis C virus (HCV) infection.
To do this, We want to compare proportion of patients with a sustained virological response when they are treated with combination interferon alpha-2b+ribavirin versus when they are treated with combination interferon alpha-2b+Placebo.

$$
\left\{
  \begin{array}{ll}
    H_0: p_1 = p_2 \\
    H_1: p_1 \ne p_2
  \end{array}
\right.
$$

With $p_1$ proportion of response in group IFN+ Ribavarin and $p_2$ proportion of response in group IFN+ Placebo.

Hypothesis given by clinician are:
- Expected response in group IFN+ Placebo is 7%.
- Expected response in group IFN+ Ribavarin is 47%
- 20% of lost to follow up.

As biostatistician, you want to determine sample size for this study with a power of 80% and confidence level of 95%.

#### Using formula

As we have a rate of lost to follow-up, we will take this into account. Lost to follow-up equal are considered as failure. So:

- proportion of response in group IFN+ Placebo is $7%*80%=0.056$

- proportion of response in group IFN+Ribavarin is $47%*80%=0.376$

- $\pi= (0.056+0.376)/2 = 0.216$

- with $\alpha=0.05$, $z_{1-\alpha /2} = 1.96$

- with $\beta=0.8$, $z_{1-\beta} = 0.84$

$N=\frac{2 \times \pi (1-\pi) (z_{1-\alpha /2} + z_{1-\beta})^2}{(p_1 - p_2)^2}$

$N=\frac{2 \times 0.216 (1-0.216) (1.96 + 0.84)^2}{(0.376 - 0.056)^2}$

$N=26$ subjects in each group.

#### Using epiDisplay
In package epiDisplay, we will use function $\text{n.for.2p}$ for estimation of sample size for two-sided proportion comparison.

```{r}
n.for.2p(p1 = 0.47 * 0.8, p2 = 0.07 * 0.8, power = 0.8, ratio = 1, alpha = 0.05)
```
$N=31$ subjects in each group.

#### Using sample simulation
```{r}

```


### OR 

For a case control study with 400 patients in each group, a proportion of 30% of event in control group, a odds ratio of 1.2 and a confidence level of 0.05 we want to :

- determine the power of this study\
- determine the good sample size to get a power of 80%.



- $n_1=400$ as number of patients in case group\
- $n_2=400$ as number of patients in control group\
- $p_2=0.3$ as proportion of event in control group\
- $p_1=\frac{1}{1+\frac{1-p_2}{OR \times p_2}}=\frac{1}{1+\frac{1-0.3}{1.2 \times 0.3}}=0.34$ as proportion of event in case group\
- $\pi=(p_1+p_2)/2 = 0.32$\
- $z_{1-\alpha/2}=1.96$\

Now we determine:

#### Using formula

- $z_{1-\beta}$

$z_{1-\beta}=\sqrt{\frac{n_1.(p_1-p_2)^2}{2\pi(1-\pi)}}-z_{1-\alpha/2} \cong \sqrt{\frac{400.(0.04)^2}{2 \times 0.32(1-0.32)}}-1.96$ \
$z_{1-\beta}= -0.76$ $\implies$ $1-\beta=0.224$\
We obtain a power of $0.224$ .

- $n_1$ and $n_2$ to have a power of $0.8$

$N=\frac{2 \times \pi (1-\pi) (z_{1-\alpha /2} + z_{1-\beta})^2}{(p_1 - p_2)^2} \cong \frac{2 \times 0.32 (1-0.32) (1.96 + 0.84)^2}{(0.34 - 0.30)^2} = 2175.04$\

To have a power of 80% we should have $2176$ patients in each group of patient.

#### Using epiDisplay

- $z_{1-\beta}$

```{r}
power.for.2p(p1=0.34,p2=0.3,n1=400,n2=400,alpha=0.05)
```
With 400 patients in each group we have a power of 20%.

- $n_1$ and $n_2$ to have a power of $0.8$

```{r}
n.for.2p(p1=1/(1+.7/.36),p2=0.3,alpha=0.05,ratio=1)
```
To have a power of 80%, we should have $n_1=n_2=2225$ patients.

#### Using sample simulation
```{r}

```


### OR 

For a case control study with 100 patients in case group, 4 control patients for one case patient, a proportion of 20% of event in control group, a odds ratio of 2 and a confidence level of 0.05 we want to :

- determine the power of this study\
- determine the good sample size to get a power of 80%.

We have:

- $n_1=100$ as number of patients in case group\
- $n_2=400$ as number of patients in control group\
- $p_2=0.2$ as proportion of event in control group\
- $p_1=\frac{1}{1+\frac{1-p_2}{OR \times p_2}}=\frac{1}{1+\frac{1-0.2}{2 \times 0.2}} \cong 0.33$ as proportion of event in case group\
- $\gamma=n_2/n_1=4$
- $\pi=\frac{p_1+\gamma p_2}{1+\gamma} = 0.227$\
- $z_{1-\alpha/2}=1.96$\

Now we determine:

#### Using formula

- $z_{1-\beta}$

$z_{1-\beta}=\frac{ \sqrt{n_1.\gamma.(p_1-p_2)}-z_{1-\alpha /2 }  \sqrt{ (1+\gamma).\pi.(1-\pi)}} {\sqrt{\gamma.p_1(1-p_1)+p_2(1-p_2)}} \cong \frac{ \sqrt{100 \times 4 \times (0.33-0.2)}- 1.96  \sqrt{ 5 \times 0.227 \times (1-0.227)}} {\sqrt{4 \times 0.33(1-0.333)+0.2(1-0.2)}}$ \
$z_{1-\beta}= 0.812$ $\implies$ $1-\beta=0.79$\
We obtain a power of $0.224$ .

- $n_1$ and $n_2$ to have a power of $0.9$

$N=\frac{(z_{1-\alpha /2 } \sqrt{(1+\gamma).\pi.(1-\pi)} + z_{1-\beta} \sqrt{\gamma . p_1(1-p_1)+p_2(1-p_2)})^2}{\gamma \times (p_1 - p_2)^2} \cong \frac{(1.96 \sqrt{(1+4).0.227.(1-0.227)} + 1.28 \sqrt{4 \times 0.33(1-0.33)+0.2(1-0.2)})^2}{4 \times (0.33 - 0.2)^2} = 2175.04$\

To have a power of 80% we should have $2176$ patients in each group of patient.

#### Using epiDisplay

- $z_{1-\beta}$

```{r}
power.for.2p(p1=0.33,p2=0.2,n1=100,n2=400,alpha=0.05)
```
With 100 patients in case and 400 in control in each group we have a power of 73.5%.

- $n_1$ and $n_2$ to have a power of $0.9$

```{r}
n.for.2p(p1=1/(1+0.8/0.4),p2=0.2,alpha=0.05,power=0.9,ratio=4)
```
To have a power of 90%, we should have 112 patients in case and 448 in control group.

#### Using sample simulation

We have:

- $p_2=0.2$ as proportion of event in control group\
- $p_1=\frac{1}{1+\frac{1-p_2}{OR \times p_2}}=\frac{1}{1+\frac{1-0.2}{2 \times 0.2}} \cong 0.33$ as proportion of event in case group\
- We will do simulation using different size of sample and select the best. From $n_1=100$ and $n_2=400$ to $n_1=200$ and $n_2=800$ with a step of 20.


```{r}


n1=seq(100,200,20) #variation of case size from 100 to 200 with a step of 20 
n2=seq(400,800,80) #variation of case size from 400 to 800 with a step of 20
size=data.frame(n1,n2) #set a frame with couple of n1 n2 as zip in python
power=rep(NA, nrow(size)) #initialize vector which receive power value of each size

for (i in 1:nrow(size)){
  set.seed(1)
  pval<-replicate(500,{
    y1=rbinom(n=size[i,1],p=1/3,size=1)
    y2=rbinom(n=size[i,2],p=0.2,size=1)
    y=rbind(table(y1),table(y2))
    chisq.test(y)$p.value
  }) #replicate will perform 500 times the random simulation of two sample with size defined and compute a chi2 test. In p val vector output we will obtain 500 p-values of chi2 test
power[i]=mean(pval<0.05) # At each i step power is the percent of p value < 0.05 over 500 simulations
}

sample_size=size[which.max(power),]

```


### Longitunal analysis

You set a longitudinal study to determine the efficacy of a new treatment vs an old treatment in survival free colon cancer. So you follow patients on 5 years. 30% of patients received. Incidence of disease is equal 2/1000 in general population. The risk is 3 more higher in population of patients. You need to determine sample size with a power of 80% and confidence level of 95%.

We have:
On 5 year:
- $p_2=1-[(1-0.002)^5] \cong 0.01$ as proportion of event in control group on 5 year\
- $p_1=RR*p_2=3*0.01 \cong 0.03$ as proportion of event in case group on 5 years\
- $\gamma=n_2/n_1=7/3$
- $\pi=\frac{p_1+\gamma p_2}{1+\gamma} =\frac{0.03+7/3*0.01}{1+7/3} \cong 0.159$\
- $z_{1-\alpha/2}=1.96$\
- $z_{1-\beta}=0.84$

#### Using formula

$N_1=\frac{(z_{1-\alpha /2 } \sqrt{(1+\gamma).\pi.(1-\pi)} + z_{1-\beta} \sqrt{\gamma . p_1(1-p_1)+p_2(1-p_2)})^2}{\gamma \times (p_1 - p_2)^2} \cong \frac{(1.96 \sqrt{(1+7/3).0.159.(1-0.159)} + 0.84 \sqrt{7/3 \times 0.03(1-0.03)+0.01(1-0.01)})^2}{7/3 \times (0.03 - 0.01)^2} = 502.96 = 503$ patients in case group\
$N_2=7/3*502.96=1173.57 = 1174$ patients in control groups\
In total we have $N=503+1174=1677$ patients to be followed on 5 years

#### Using epiDisplay

```{r}
n.for.2p(p1=0.03, p2=0.01, alpha=0.05, power=0.8, ratio=7/3)
```
With this method, we have $N_1=571$ patients and $N_2=1332$ patients follow on five years.

#### Using sample simulation
```{r}

```
